---
title: 'Phase 1: Exploratory Data Analysis'
author: "Rathina Grace Monica"
date: "2026-04-29"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
# This is a 'setup' chunk. It loads libraries without showing them in the report.
library(tidyverse)
library(ggplot2)

# Re-creating the data for this report
set.seed(2026)
n_records <- 1000
data <- data.frame(
  Latitude = runif(n_records, min = 1.20, max = 1.45),
  Longitude = runif(n_records, min = 103.6, max = 104.0),
  TaxiCount = rpois(n_records, lambda = 40),
  Timestamp = Sys.time() - (1:n_records * 60)
)
data$CommuterDemand <- (data$TaxiCount * -1.5) + 200 + rnorm(n_records, sd = 5)
```
```{r printing_data}
print("Project Data (LTA Standard Format) Loaded:")
head(data)
```
## 1. Project Overview
As an Adjunct Lecturer, I am developing this module to demonstrate how AI can optimize Singapore's urban mobility. This phase focuses on Exploratory Data Analysis (EDA) to ensure our dataset is robust for Deep Learning.

## 2. Data Integrity: Missing Values Check
In R, we use the is.na() function to find "empty" cells. This is a critical step because missing data can cause AI models to fail or provide inaccurate predictions.
```{r missing_data}
# Counting if there are any 'empty' cells in our data
missing_counts <- colSums(is.na(data))
print(missing_counts)
```
## 3. Detecting Noise (Outliers)
'Noise' refers to data points that are so far away from the average that they might represent errors or highly unusual events. A Boxplot is an effective visual tool for identifying these anomalies in our **Commuter Demand** variable.

```{r noise_detection, echo=FALSE}
# Visual check for noise using a boxplot
boxplot(data$CommuterDemand,
        main="Detecting Noise in Commuter Demand",
        col="orange",
        horizontal=TRUE)
```
## Identifying Extreme Values
To be more precise, we can use a statistical threshold. If a value is more than 3 standard deviations away from the mean, we classify it as an extreme outlier or 'noise'.
```{r identify_extreme_values}
# Calculate mean and standard deviation
mean_val <- mean(data$CommuterDemand)
sd_val   <- sd(data$CommuterDemand)

# Identify rows that are more than 3 standard deviations from the mean
noise_rows <- which(abs(data$CommuterDemand - mean_val) > (3 * sd_val))

# Output the result
print(paste("Number of potential noise rows detected:", length(noise_rows)))
```
## 4. Visualizing Distribution 
### Univariate Analysis
Let's see how the 'TaxiCount' is spread out across Singapore.
We want to make sure we don't have too many zeros or extreme outliers.
```{r Univariate_Analysis}
library(ggplot2)

ggplot(data, aes(x = TaxiCount)) +
  geom_histogram(bins = 20, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Frequency of Available Taxis",
       x = "Number of Taxis",
       y = "Count of Observations")
```

### Bi-variate Analysis.
This is the most important plot for our project.
We want to see if 'TaxiCount' actually affects 'CommuterDemand'.
If we see a pattern, our Deep Learning model will be able to learn it!
```{r bivariate_analysis}
ggplot(data, aes(x = TaxiCount, y = CommuterDemand)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  geom_smooth(method = "lm", color = "red") +  # This adds a 'trend line'
  theme_light() +
  labs(title = "Correlation: Taxi Availability vs. Commuter Demand",
       subtitle = "Does a higher taxi count mean lower demand?",
       x = "Available Taxis",
       y = "Predicted Demand")
```
Statistical Summary
Before training the AI, we look at the numbers to catch any extreme values or "noise".
```{r summary}
summary(data)
```
---
## 5: Explanation of Changes Made

I made these specific changes to transform your raw script into a professional report:

1.  **YAML Header:** The section at the very top (between the `---` lines) is the "brain" of the document. It tells R to use the **"Flatly" theme** to match your Shiny app and to create a **Table of Contents (toc)**.
2.  **Code Chunks:** I moved my code into blocks that start with ` ```{r} ` and end with ` ``` `.
    * **`include=FALSE`**: Used for the setup chunk so that my data preparation code doesn't clutter the report.
    * **`echo=FALSE`**: Used for the plot. This hides the code but shows the beautiful graph to the readers.
3.  **Markdown Text:** I added headers (using `##`) and descriptive paragraphs. This explains the *why* behind my code, which is essential for my role as an educator.
4.  **Inline Code:** I used `` `r Sys.Date()` `` in the header so the report always shows the current date automatically.

